\bf On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2025-03-04 Manuscript/manuscript.tex
opening paragraphs of sections 2 and 3; additional smoothing

%Needs to be looked at
The rest of this paper is organized as follows. Section~\ref{sec:problem1}
investigates RTs of athletes who competed at the 2022 World Championships to
examine differences between other competitions.  Section~\ref{sec:problem2}
investigates RTs of athletes from 1999 to 2023 to determine a reaction barrier
ground in statistical analysis.  Within each of the above sections, the data,
methods used, and the results are presented.
Finally, Section~\ref{sec:concludingremarks} highlights the
paper’s impact, limitations, and potential for future research.
All data and code for our analysis are provided in the Supplementary
Materials.

\section{Analysis of the 2022 World Championships RTs}
\label{sec:problem1}

@@

To investigate whether RTs at the 2022 World Championships
were significantly different from other competitions, we
used data from male athletes in the 110-meter hurdles and 100-meter
dash, and female athletes in the 100-meter hurdles and 100-meter dash,
provided they competed in the 2022 World Championships and at least
one other competition (2022 national championships, 2019 World
Championships, or 2023 World Championships).  We excluded data from
200-meter dashes and longer events, as their RTs were found
to be significantly different. Negative RTs were excluded from the analysis,
while positive disqualified RTs were included due to their low frequency and not
being obvious outliers.

This study focuses on RTs from sprint events where reaction
time plays a critical role. Specifically, we included data from men’s
100-meter dashes and 110-meter hurdles, as well as women’s 100-meter
dashes and 100-meter hurdles. Statistical tests showed no significant
differences in RTs between these events, supporting their
inclusion in a unified analysis. However, we excluded data from
200-meter dashes and longer events, as their RTs were found
to be significantly different. This difference is not unexpected, as
RTs are generally less critical in longer sprint events
compared to shorter distances where explosive starts are paramount.
Negative RTs were excluded from the analysis, while positive
disqualified RTs were included due to their low frequency and not
being obvious outliers. % A
% sensitivity analysis regarding the inclusion of positive disqualified
% RTs is provided in the Supplementary Material.

@@

To assess the abnormality of the 2022 World Championships, we examined
data beyond World Championship events. This decision was motivated by
an exploratory analysis of how United States (US) athletes performed
at the 2022 US Track and Field Championships, held from June 23--26,
2022, at Hayward Field in Eugene, Oregon. Since this venue also hosted
the 2022 World Championships in August, it
provided a unique opportunity to compare. There were four US
110-meter hurdle athletes—Trey Cunningham, Daniel Roberts,
Grant Holloway, and Devon Allen—who competed in both events. All four
athletes recorded faster RTs in every World Championships race
compared to their performances at the national-level event. Similarly,
all four US 100-meter dash athletes---Marvin Bracy, Fred Kerley,
Travyon Bromwell, and Christian Coleman---also recorded slower RTs at
the US Championships than at the World Championships.
To expand the dataset, we included RTs from 100-meter hurdle
and 100-meter dash athletes who competed in other national
competitions held between May and July 2022 across various
countries. Compiling this data presented challenges, as RTs were not
centrally archived and often required searching country-specific
websites, with many results recorded in native languages.

The final dataset consisted of RTs from athletes who competed
in both national and international competitions. RTs from
preliminary heats, semifinals, and finals were included to ensure
sufficient data for analysis. Excluding preliminary heats would have
significantly reduced the number of athletes and clusters. Each
athlete was considered as a cluster, with observed RTs
from both the `treatment' group (2022 World Championships) and
the `control' group (national competition) within the cluster. Cluster
sizes ranged from three to six, with a median size of four. Because gender
is known to influence RTs \citep{babicc2009reaction,
  lipps2011implications}, we prepared data for men and women
separately, resulting in 80 RTs from 17 athletes for each
gender. This setup forms the basis for the rank-based comparison for
clustered data with subunit-level grouping, which accounts for the
within-athlete dependence inherent in this data structure. The top
panel of Figure~\ref{fig:RankScatterplots} shows the RTs
of those who competed both at national competitions and the 2022 World
Championships.

@@

To compare RTs at the 2022 World Championships with those
recorded in the 2019 and 2023 World Championships, we prepared datasets
including RTs from athletes who competed in the 2022 World
Championships and at least one of the 2019 or 2023 World Championships.
Specifically, we identified athletes who participated in both the 2019
and 2022 World Championships, as well as those who participated in both
the 2022 and 2023 World Championships. In the 2019-2022 comparison,
RTs from 2022 were treated as the `treatment' group, with
2019 serving as the `control' group. Similarly, in the 2022-2023
comparison, RTs from 2022 were treated as the `treatment'
group, with 2023 serving as the `control' group. This structure allowed
us to prepare datasets suitable for examining RTs of athletes
who competed across multiple World Championships.

@@

Figure~\ref{fig:RankScatterplots} shows the RTs of athletes
who competed in both the 2019 and 2022 World Championships (middle
panel) and those who competed in both the 2022 and 2023 World
Championships (lower panel). The athletes included in the 2022 data
differ between these two comparisons, as the set of athletes who
competed in both 2019 and 2022 is not the same as the set who competed
in both 2022 and 2023. To be included, athletes must have competed in
at least one race at each championship. Notably, Devon Allen recorded
the fastest RTs in both the Finals and Semifinals of the
2022 World Championships, but his disqualification was determined by a
difference of just 0.002 seconds, with RTs of 0.101 and
0.099 seconds, respectively. This highlights the critical role of
RT precision in elite-level competition.

@@

To test the conjecture that the 2022 World Championships timing device may have
led to faster recorded RTs, we compare the RTs of the same
athletes who have attended both the 2022 World Championships and other
competitions.
In this setting, we have clustered data with subunit grouping. In particular,
each athlete is a cluster and the multiple RTs from the same athlete
can be from either the 2022 World Championships or otherwise.
Let $X_{ij}$ be the $j$th RT of athlete~$i$, $i = 1, \ldots, n$,
$j = 1, \ldots, m_i$ where $m_i$ is the number of observations from
athlete~$i$. Let $\delta_{ij}$ be the group indicator of $X_{ij}$; $\delta_{ij}
= 1$ if $X_{ij}$ is in group~1 (2022 World Championships) and $\delta_{ij} = 0$
otherwise. Athletes are
assumed to be independent, while subunit observations from the same athlete are
not. The null hypothesis $H_0$ to be tested is that there is no difference
between the two groups; i.e., the distribution of $X_{ij}$ remains the same
regardless of the group indicator $\delta_{ij}$.

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subunit-level grouping. The test is designed based on a
within-cluster resampling principle. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the pseudo-sample.
The test statistic $S$ is the average of $W^*$, averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a $Z$ statistic which follows a standard normal distribution
asymptotically \citep[p.910]{datta2005rank}.

When the sample size is small, the asymptotic normal distribution may
not be reliable, so we also use 1~million random permutations to
simulate the null distribution of the test statistic.
This method is available from the \texttt{clusWilcox.test()} function
with \texttt{method = `ds'} (for \underline{D}atta and \underline{S}atten) and
\texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

@@

\section{Analysis of the 0.1 RT Barrier}
\label{sec:problem2}

@@

\section{Discussion}\label{sec:concludingremarks}

>>>>

%Needs to be looked at
The rest of this paper is organized as follows. Section~\ref{sec:problem1}
investigates RTs of athletes who competed at the 2022 World Championships to
examine differences between other competitions.  Section~\ref{sec:problem2}
investigates RTs of athletes from 1999 to 2023 to determine a reaction barrier
ground in statistical analysis.  Within each of the above sections, the data,
methods used, and the results are presented.
Finally, Section~\ref{sec:disc} highlights the
paper’s impact, limitations, and potential for future research.
All data and code for our analysis are provided in the Supplementary
Materials.

\section{Assessing the 2022 World Championships RTs}
\label{sec:problem1}

The notable faster RTs at the 2022 World Championships
raised concerns about systematic biases. 
For example,  the median RTs across multiple sprint events were found
to be the lowest in recent history; the number of RTs recorded under
0.115 at men’s 100m dash and 110m hurdles was 25, much greater than that in
2019, which was 3 \citep{johnson2022was}. This striking disparity
suggests a potential systematic difference in RT measurements at the
2022 World Championships, necessitating a formal statistical
investigation. We assess whether RTs at the 2022 World Championships
were systematically faster by comparing them against (1) national
competitions in 2022 and (2) RTs from the 2019 and 2023 World
Championships to evaluate deviations from historical norms, using data
from athletes who competed in both events in the comparison.

@@

To investigate whether RTs at the 2022 World Championships
were significantly different from other competitions, we
used data from male athletes in the 110-meter hurdles and 100-meter
dash, and female athletes in the 100-meter hurdles and 100-meter dash,
provided they competed in the 2022 World Championships and at least
one other competition (2022 national championships, 2019 World
Championships, or 2023 World Championships). Statistical tests showed no
significant differences in RTs between these events, supporting their
inclusion in a unified analysis. However, we excluded data from
200-meter dashes and longer events, as their RTs were found
to be significantly different. This exclusion is expected, as reaction
time plays a smaller role in longer sprint events, where acceleration
off the blocks is less decisive.
Negative RTs were excluded from the analysis, while positive disqualified RTs
were included due to their low frequency and not being obvious outliers.

@@

We first compare RTs from the 2022 World Championships to national
competitions earlier that year to assess whether the same athletes
reacted differently across events.
Prior to a formal analysis, we examined how United States (US) athletes
performed at the 2022 US Track and Field Championships, held from June 23--26,
2022, at Hayward Field in Eugene, Oregon. Since this venue also hosted
the 2022 World Championships in August, it
provided a unique opportunity to compare. Among the four US
110-meter hurdle athletes---Trey Cunningham, Daniel Roberts,
Grant Holloway, and Devon Allen---all
recorded faster RTs in every World Championships race
compared to their performances at the national-level event. Similarly,
all four US 100-meter dash athletes---Marvin Bracy, Fred Kerley,
Travyon Bromwell, and Christian Coleman---also recorded faster RT's
at the World Championships, reinforcing concerns 
about systematic differences in reaction measurements.

To expand the dataset, we included RTs from 100-meter hurdle
and 100-meter dash athletes who competed in other national
competitions held between May and July 2022 across various
countries. Compiling this data presented challenges, as RTs were not
centrally archived and often required searching country-specific
websites, with many results recorded in native languages.

The final dataset consisted of RTs from athletes who competed
in both national and international competitions. RTs from
preliminary heats, semifinals, and finals were included to ensure
sufficient data for analysis. Excluding preliminary heats would have
significantly reduced the number of athletes and clusters. Each
athlete was considered as a cluster, with observed RTs
from both the `treatment' group (2022 World Championships) and
the `control' group (national competition) within the cluster. Cluster
sizes ranged from three to six, with a median size of four. Because gender
is known to influence RTs \citep{babicc2009reaction,
  lipps2011implications}, we prepared data for men and women
separately, resulting in 80 RTs from 17 athletes for each
gender. This structure enables a rank-based comparison for clustered 
data, which properly accounts for the within-athlete dependence inherent 
in this dataset. The top
panel of Figure~\ref{fig:RankScatterplots} shows the RTs
of those who competed both at national competitions and the 2022 World
Championships.

@@

To determine whether RTs at the 2022 World Championships were
anomalous within the World Championships series, we compare them to
RTs from the 2019 and 2023 events. In the 2019-2022 comparison,
RTs from 2022 were treated as the `treatment' group, with
2019 serving as the `control' group. Similarly, in the 2022-2023
comparison, RTs from 2022 were treated as the `treatment'
group, with 2023 serving as the `control' group. This structure allowed
us to prepare datasets suitable for examining RTs of athletes
who competed across multiple World Championships.

@@

Figure~\ref{fig:RankScatterplots} shows the RTs of athletes
who competed in both the 2019 and 2022 World Championships (middle
panel) and those who competed in both the 2022 and 2023 World
Championships (lower panel). Since the sets of athletes differ between 
the two comparisons, each analysis provides independent evidence 
for evaluating potential anomalies in 2022 RTs. Notably, Devon Allen
recorded the fastest RTs in both the Finals and Semifinals of the
2022 World Championships, but his disqualification was determined by a
difference of just 0.002 seconds, with RTs of 0.101 and
0.099 seconds, respectively. This highlights the critical role of
RT precision in elite-level competition.

@@

To test whether the timing system at the 2022 World Championships 
systematically produced faster RTs, we compare RTs recorded at this
event against those from the same athletes in other competitions. In
this setting, we have clustered data with subunit grouping. Each
athlete serves as a cluster, with multiple RTs recorded from the same
athlete across different competitions.
Let $X_{ij}$ be the $j$th RT of athlete~$i$, $i = 1, \ldots, n$,
$j = 1, \ldots, m_i$ where $m_i$ is the number of observations from
athlete~$i$. Let $\delta_{ij}$ be the group indicator of $X_{ij}$; $\delta_{ij}
= 1$ if $X_{ij}$ is in group~1 (2022 World Championships) and $\delta_{ij} = 0$
otherwise. Athletes are
assumed to be independent, while subunit observations from the same athlete are
not. The null hypothesis $H_0$ to be tested is that there is no difference
between the two groups; i.e., the distribution of $X_{ij}$ remains the same
regardless of the group indicator $\delta_{ij}$.

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subunit-level grouping. The test is based on a
within-cluster resampling approach that preserves the within-cluster
dependence. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the pseudo-sample.
The test statistic $S$ is the average of $W^*$, averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a $Z$ statistic which follows a standard normal distribution
asymptotically \citep[p.910]{datta2005rank}.

For small sample sizes, the asymptotic normality may be unreliable. To
address this, we also use 1~million random permutations to
simulate the null distribution of the test statistic.
This method is available from the \texttt{clusWilcox.test()} function
with \texttt{method = `ds'} (for \underline{D}atta and \underline{S}atten) and
\texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

@@

\section{Evaluating The 0.1 Second RT Threshold}
\label{sec:problem2}

This section evaluates whether the 0.1-second RT threshold remains a valid 
criterion for disqualification in elite sprinting. Using historical RT data from 
World Championships since 1999, we fit a generalized Gamma model with 
venue- and heat-level random effects to estimate the probability of RTs below 
0.1 seconds. This statistical framework allows us to assess whether an 
alternative threshold would better align with observed reaction time 
distributions.

@@

\section{Discussion}\label{sec:disc}
